Formalizing Stack Safety as a Security Property

### 2023-01-31 submission5/ideas-v2.tex
comments

Security properties tend to be conservative, identifying violations that would be
harmless in practice. But enforcement mechanisms make trade-offs in the name of performance
and of compatibility with existing codebases. In a crucial motivating example for this work,
Roessler and DeHon \cite{} present a collection of tag-based enforcement mechanisms
using PIPE \cite{???}, a tagged hardware reference monitor. In monitors such as PIPE and
STAR \cite{???}, every value in memory and registers is paired with a {\em metadata tag}.
A set of software-defined rules, termed a micro-policy, determines the tags for operation
results based on the tags of their inputs---or, if the input tags are in a dangerous
combination, it may halt the machine (a ``failstop.'')

Roessler and DeHon find that the intuitive way to enforce stack safety in this setting
is inefficient when allocating and clearing stack frames, because each word must be initialized
with a tag corresponding to the (depth of the) current function activation. They propose
{\em Lazy Tagging and Clearing}, in which tags are not initialized at allocation, but only
when a function writes to memory. This means that writes cannot be prevented, necessarily
violating existing models of stack safety. Instead the micro-policy enforces that memory be
read by the same fuction that wrote it. The owner of the data will never see the changes,
so it is secure in the sense that the attacker is unsuccessful at changing program behavior
(except, perhaps, by causing a fault.) We elaborate that sense, and test whether the
micro-policy fulfills it.

@@

@@

\subsubsection{Low-Level Component Support}
%
Our framework considers a \emph{single-, low-level view} of all program code and
data; a key advantage of this approach is that it allows us to remain
\emph{language-agnostic}.
%
Some software-based protection mechanisms are designed specifically with
a \emph{high-level language} in mind, e.g. bounds checking and split stacks.
They tend to offer stronger security guarantees than other software techniques.
In exchange, their threat model becomes weaker, as they make additional
assumptions about the admissible behavior of program components
%
(for example, bounds checking only applies to code that has been instrumented by
a \emph{secure compiler} \ref{}).
%
In general, a multi-level extension of the framework offering adequate reasoning
capabilities for this class of policies would need to model multiple source
languages and a notion of code provenance---a reflection of the call stack as the
melting pot of language abstractions.
%
The common case of one source and one target language mediated by a compiler may
be solved by a simpler lifting of our framework from the machine to the source
level \rb{e.g., SoftBound could use a C language semantics and Tagged C for
security modeling (lifted properties would likely hold under the expected
assumptions)}.

@@

The properties form a hierarchy, shown in [TODO: diagram].
Within properties that are orthogonal
some are likely to take priority in terms of their
importance to security. Almost all stack-safety enforcement mechanisms will try to enforce \(WBCF\),
as control-flow attacks such as ROP are a major vulnerability that can be targeted via the stack.
Many also cover some degree of (caller) integrity. Caller confidentiality has been fairly neglected,
and callee-oriented properties even moreso.

@@

The set of security-relevant operations (\(\Psi\)) covered in this paper is given in
\cref{tab:psi}. Those that we explore in detail in \cref{sec:example}
are shown in white: calls, returns, and the allocation and deallocation of local memory.
In \cref{sec:machine}), we cover additional operations that add parameters to the examples',
to support various forms of memory sharing \ifexceptions , exceptions, \fi
and tail-call elimination. These are shown in light gray.
We give a proof-of-concept formalization of a more sophisticated, heap-like sharing model
based on pointer provenance in \cref{app:ptr}, represented by the dark grey
operations, but we do not test this.

@@

\sna{Might like a better name than object, because morally it ought to apply to arguments as well.}

>>>>

Security properties tend to be conservative, identifying violations that may be
harmless in practice. But enforcement mechanisms make trade-offs in the name of performance
and of compatibility with existing codebases. In a crucial motivating example for this work,
Roessler and DeHon \cite{} present a collection of tag-based enforcement mechanisms
using PIPE \cite{???}, a tagged hardware reference monitor. In monitors such as PIPE and
STAR \cite{???}, \apt{double-check this} every value in memory and registers is paired with a {\em metadata tag}.
A set of software-defined rules, termed a micro-policy, determines the tags for operation
results based on the tags of their inputs---or, if the input tags are in a dangerous
combination, it may halt the machine (a ``failstop.'')

The most obvious way to enforce stack safety in this setting is to tag each word
in the stack frame with the (depth of) its owning function activation: tags are
set when the frame is allocated, checked against the current activation tag at each read or write,
and cleared when the frame is deallocated. 
However, Roessler and DeHon find that this approach is inefficient, because tag setting
and clearing must be performed even on elements of the frame that are never accessed,
which is quite common, e.g. in code that allocates unnecessarily large local arrays.
They therefore propose
{\em Lazy Tagging and Clearing}, in which tags are not initialized at allocation, but only
when a function writes to memory. This means that writes cannot be prevented, necessarily
violating existing models of stack safety.\apt{Like what??}  Instead the micro-policy enforces that memory be
read by the same function activation that wrote it. The owner of the data will never see the changes,
so it is secure in the sense that the attacker is unsuccessful at changing program behavior
(except, perhaps, by causing a fault). We elaborate this notion of observational safety, and test whether the
micro-policy fulfills it.

@@

\apt{Sorry, but I don't understand what this section is trying to achieve.
What is the ``comparison'' in Table~\cite{tab:relwork} for?
It would be great if we could describe how our properties are suitable
for describing the kinds of things various protection mechanisms are trying
to acheive, and I can see a sketch of this going in related/future work, but 
in fact we DON'T have any evidence for anything except micropolicies.}

@@

\subsubsection{Low-Level Component\apt{``Code''?} Support}
%
Our framework considers a \emph{single-, low-level view} of all program code and
data; a key advantage of this approach is that it allows us to remain
\emph{language-agnostic}.
%
Some software-based protection mechanisms are designed specifically with
a \emph{high-level language} in mind, e.g. bounds checking and split stacks.
They tend to offer stronger security guarantees than other software techniques.
In exchange, their threat model becomes weaker, as they make additional
assumptions about the admissible behavior of program components
%
(for example, bounds checking only applies to code that has been instrumented by
a \emph{secure compiler} \ref{}).
%
In general, a multi-level extension of the framework offering adequate reasoning
capabilities for this class of policies would need to model multiple source
languages and a notion of code provenance---a reflection of the call stack as the
melting pot of language abstractions.
%
The common case of one source and one target language mediated by a compiler may
be solved by a simpler lifting of our framework from the machine to the source
level \rb{e.g., SoftBound could use a C language semantics and Tagged C for
security modeling (lifted properties would likely hold under the expected
assumptions)}.

@@

The properties form a hierarchy, shown in [TODO: diagram].
Within properties that are orthogonal
some are likely to take priority in terms of their
importance to security. Almost all stack-safety enforcement mechanisms will try to enforce {\wbcf},
as control-flow attacks such as ROP are a major vulnerability that can be targeted via the stack.
Many also cover some degree of (caller) integrity. Caller confidentiality has been fairly neglected,
and callee-oriented properties even more so.

@@

The set of security-relevant operations (\(\Psi\)) covered in this paper is given in
\cref{tab:psi}. Those that we explore in detail in \cref{sec:example}
are shown in white: calls, returns, and the allocation and deallocation of local memory.
In \cref{sec:machine}), we cover additional operations that add parameters to the examples,
to support various forms of memory sharing \ifexceptions , exceptions, \fi
and tail-call elimination. These are shown in light gray.
We give a proof-of-concept formalization of a more sophisticated, heap-like sharing model
based on pointer provenance in \cref{app:ptr}, represented by the dark grey
operations, but we do not test this.

@@

\sna{Might like a better name than object, because morally it ought to apply to arguments as well.\apt{Maybe ``active'' ?}}

### 2023-01-31 submission5/intro-v2.tex
comments

Computer programming at any scale relies on \emph{functions} (variously known as
subroutines, procedures, methods, etc.) as a \emph{foundational abstraction} to
structure programs into building blocks representing units of computation that
can \emph{call} one another to create larger computations.
%
At a low level, a function call comprises its own data (private or shared), as
well as information about its \emph{caller} and \emph{callee} functions. The
runtime must keep track of the local contexts of all function activations
at any given time; the \emph{fundamental data structure} charged with realizing
this abstraction is the \emph{(call) stack}.
%
\rb{This should be well known, so no need for illustration?}

@@

The correct execution on a program---and its security!---is contingent on all of
its functions properly implementing this abstraction by means of a careful use
of the call stack. Be it accidental or intentional, incorrect management of the
stack is at the root of a myriad of faults whose effects range from (rarely)
innocuous to (usually) catastrophic.
%
From a security perspective, attacks on the stack (say, from a malicious
function that chooses to abuse its access to the contents of this data
structure) are as fundamental, varied and pervasive as the functional
abstraction itself.
%
\rb{Begins to prefigure the idea of multiple stack implementations.}

@@

Indeed, the call stack is a perennial target for low-level attacks, leading to
dire consequences, from leakage and corruption of private stack data
to control-flow hijacking.
%
Even a cursory glance at the most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read), \#7 (CWE-416, Use
After Free) or \#11 (CWE-476, NULL Pointer Dereference).
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \rb{cf. Eternal war in memory, Google/Microsoft
memory safety statistics, etc.}.
%
To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}---we discuss these in more
detail in \cref{XXX}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
\rb{Broad enough? Do we want to add more techniques?}
%
Yet despite this wealth of potential solutions, the stack remains a central
security concern---why?

Security mechanisms designed to prevent a certain attack can be brittle---successfully
eliminating the attack as it exists at the time of publication, but leaving room for attackers
to find new, similar attacks.
%
%% \rb{And so the cat-and-mouse game continues.}
%
To mitigate this risk, \rb{we claim that} mechanisms should aim to enforce \emph{formal
properties} of safe behavior that can be proven or tested. Such properties become the
\emph{specification} against which enforcement can be validated; even enforcement mechanisms that do
not fulfill a property benefit from the ability to articulate \emph{why} they fail.
%
%% \rb{Good stuff! There is some overlap between this discussion and the opening
%%   motivation, maybe we could restructure things a bit so that those threads are
%%   unified?}
%
This is the problem we begin to address in this paper \rb{or a few more such
  connecting words}.

@@

@@

\paragraph{Micro-Policies}
%
\rb{(This partly overlaps with the technical preliminaries, try to offer a quick
  synthesis here.)}
%
Micro-policy machines are a family of tagged architectures that decorate
addressable memory and registers with arbitrary metadata tags, and extend the
processor with a programmable hardware monitor that uses custom rules on
metadata tags to implement many different kinds of security policies.
%
This flexibility can be harnessed to implement a wide array of stack protection policies with relative ease, covering and going beyond the capabilities \rb{word choice?} of the various software-based techniques \cite{} \rb{as demonstrated by}.
%
These policies comprise very general forms of \emph{lazy enforcement}, which can tolerate
potential but harmless violations \rb{the system programmer's motto: ``trust me,
  I know what I'm doing''} while still detecting them if and when their effects
become tangible.
%
Micro-policies are one of the centerpieces of our technical development; we
present them in more detail in \cref{XXX}.

@@

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. We consider the criteria in (3) to represent universal definitions
that are parameterized by (1).
%
\rb{FIXME: Contrast some universal definitions (or a parametric framework) with
  the concrete instance we study, again with its interest properly motivated.
  (Coming in key ideas?) No ``gold standard''?}

>>>>

Computer programming % at any scale
relies on \emph{functions} (variously known as
subroutines, procedures, methods, etc.) as a foundational abstraction.
\apt{General comment: I think it's better to reserve italics for terms being defined, and avoid using them just for rhetorical emphasis.}
Functions are units of computation that
can \emph{call} one another to realize larger computations.
%
At a low level, a function call \apt{``activation''?} comprises its own data (private or shared), as
well as information about its \emph{caller} and \emph{callee} functions. The
runtime must keep track of the local contexts \apt{not defined} of all function activations
at any given time. The fundamental data structure charged with realizing
this abstraction is the \emph{(call) stack}.
%
\rb{This should be well known, so no need for illustration?}

@@

The correct execution of a program---and its security!---is contingent on all of
its functions properly implementing this abstraction by means of a careful use
of the call stack. Be it accidental or intentional, incorrect management of the
stack is at the root of a myriad of faults whose effects range from % (rarely)
innocuous to % (usually)
catastrophic.
%
From a security perspective, attacks on the stack (say, from a malicious
function that chooses to abuse its access to the contents of this data
structure) are as fundamental, varied and pervasive as the functional
abstraction itself
%
\rb{Begins to prefigure the idea of multiple stack implementations.}
\apt{I think this is all well-known.}

@@

Indeed, the call stack is a perennial target for low-level attacks, leading to
dire consequences, from leakage and corruption of private stack data
to control-flow hijacking.
%
Even a cursory glance at the most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read), \#7 (CWE-416, Use
After Free) or \#11 (CWE-476, NULL Pointer Dereference).\apt{That last one is not usually a stack issue.}
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \rb{cf. Eternal war in memory, Google/Microsoft
memory safety statistics, etc.}.
\apt{OK to say, but this is well-known.}
%
To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}---we discuss these in more
detail in \cref{XXX}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
\rb{Broad enough? Do we want to add more techniques?}
%
Yet despite this wealth of potential solutions, the stack remains a central
security concern---why?
\apt{Mainly because preventing common stack problems generally causes a performance hit.
  And, to a (much) lesser extent, because C/C++ don't carry length information with
  arrays.  The sequencing here seems to suggest that we are going to \emph{solve} the
  problem of enforcing stack safety by defining it better---which I don't buy at all!}

Security mechanisms designed to prevent a certain attack can be brittle---successfully
eliminating the attack as it exists at the time of publication, but leaving room for attackers
to find new, similar attacks.
%
%% \rb{And so the cat-and-mouse game continues.}
%
To mitigate this risk, \rb{we claim that} mechanisms should aim to enforce \emph{formal
properties} of safe behavior that can be proven or tested. Such properties become the
\emph{specification} against which enforcement can be validated; even enforcement mechanisms that do
not fulfill a property benefit from the ability to articulate \emph{why} they fail.

%
%% \rb{Good stuff! There is some overlap between this discussion and the opening
%%   motivation, maybe we could restructure things a bit so that those threads are
%%   unified?}
%

@@

\apt{This section was a nice summary of enforcement approaches, but I am not
  yet seeing why it justifies so much paper real estate, especially if we only 
  actually apply our property definitions to (lazy) micropolicies. And in any case,
  it makes for a huge delay before getting to our work and contributions.}

@@

\paragraph{Micro-Policies}
%
\rb{(This partly overlaps with the technical preliminaries, try to offer a quick
  synthesis here.)}
%
Micro-policy machines are a family of tagged architectures that decorate
addressable memory and registers with arbitrary metadata tags, and extend the
processor with a programmable hardware monitor that uses custom rules on
metadata tags to implement many different kinds of security policies.
%
This flexibility can be harnessed to implement a wide array of stack protection policies with relative ease, covering and going beyond the capabilities \rb{word choice?} of the various software-based techniques \cite{} \rb{as demonstrated by}.
%
These policies comprise very general forms of \emph{lazy enforcement}, which can tolerate
potential but harmless violations \rb{the system programmer's motto: ``trust me,
  I know what I'm doing''} while still detecting them if and when their effects
become observable. 
%
Micro-policies are one of the centerpieces of our technical development; we
present them in more detail in \cref{XXX}.

@@

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. We consider the criteria in (3) to represent universal definitions
that are parameterized by (1).
%
\rb{FIXME: Contrast some universal definitions (or a parametric framework) with
  the concrete instance we study, again with its interest properly motivated.
  (Coming in key ideas?) No ``gold standard''?}
\apt{Yes, there are at least two different kinds of parameterization going on.
  One has to do with which properties we want to, or are able to, enforce; e.g.,
  for a language with tail calls or exceptions, we need a more complex definition of the WBCF property itself.
  Another has to do with how standard features, like argument passing, are implemented
  by particular hardware/ABIs -- which shouldn't show up in the property definition.
  We need to be clearer about this.}
